---
title: "Making the CuteCats package"
author: "Sean Davis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making the CuteCats package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Why an R package?

Organizing code, data, and thoughts (via documentation and vignettes) for your
own analysis does not necessarily require an R package. However, you may want
to share your materials, analysis, data, and functions with your future self
or others. Hadley Wickham envisions an R package 
[as the easiest way to do so](ttp://adv-r.had.co.nz/Philosophy.html).

R packages contain provenance (who wrote it, when, etc.), data, R code, 
documentation, tests, and really almost any other files important to the 
functionality of the package. Packages are installable, reusable, shareable,
versioned, and extensible. 

It is not about making a beautiful, perfect R package. The idea is about 
creating a bare-minimum R package so that you don’t have to keep thinking 
to yourself, “I really should just make an R package with these functions 
so I don’t have to keep copy/pasting them.” It isn’t all about sharing
sharing your code, although it is easy to do once you have an R package. It 
is about saving yourself time and serving as an external brain for your
code and analysis. 

# What is an R package?

The simplest R package is a directory with a file called `DESCRIPTION` (all
caps, no suffix). In addition, a package usually has an `R` directory that
contains R functions in files. Additional, optional folders and files serve 
other purposes. A `data` folder allows for data of specific types
of files (csv, rda). The `inst` directory will be added to the package when
it is installed; this is a good place to add supplementary files that are
needed for a package but that don't fit in the `data` folder. 

The files in a cute little *PANDA* package might look like:

```
├── DESCRIPTION
├── NAMESPACE
├── R
│   ├── cute_panda.R
│   └── eucalyptus.R
├── abc.Rproj
├── data
│   └── panda.csv
└── inst
    └── ext_images
        ├── baby_panda.png
        └── mommy_panda.png
```

# How do I create an R package?

## Step 0: Install needed pacakges

```{r eval=FALSE}
BiocManager::install("devtools")
BiocManager::install("roxygen2")
```

## Step 1: Create a skeleton R package

The `devtools` package contains lots of helpers for R package development.
The first function from that package we use is the `create` function:

```{r eval=FALSE}
devtools::create("CuteCats")
```

After executing, the `CuteCats` directory will contain a few files new files:

```
├── CuteCats.Rproj 
├── DESCRIPTION
├── NAMESPACE
└── R
```

## Step 2: Edit the DESCRIPTION file

The `DESCRIPTION` file looks something like:

```
Package: CuteCats
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person("First", "Last", , "first.last@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.0
```

Edit this file according to the short documentation. It is just a text file,
so edit it in RStudio works just fine.

## Step 3: Add functions

We can include functions in our package by writing them and then saving them
in the `R` directory. A file can contain one or more functions. The names
of the files don't matter much, but the file must end with `.R`. 

For example, open a new R script in RStudio and create the `love_cats()`.

```{r}
love_cats <- function(love=TRUE){
    if(love==TRUE){
        print("I love cats!")
        return(TRUE)
    }
    else {
        print("I am not a cool person.")
        return(FALSE)
    }
}
```

For our second function, let's take advantage of the presence of the cute
kitten that graced us with its presence during the CSHL course in 2022. This
function is going to display a photo of the Hooper House kitten. Create a 
new R script file in RStudio and add the `hooper_kitten` to it.

```{r}
hooper_kitten <- function() {
    file_loc = system.file('images/hooper_kitten.png', package='CuteCats')
    pic = png::readPNG(file_loc)
    grid::grid.raster(img)
}
```

Save the function as `hooper.R` in the `R` directory.

Note that this function requires two new packages, the `png` package and the 
`grid` package. We tell R that our package needs these two packages by adding
the following to our `DESCRIPTION` file below the `License` line (see above).

```
Imports: png, grid
```

By doing so, when people install our R package, the `png` and `grid` packages
will also be installed if needed. 

## Step 4: Add data and other files

Lets add our cat picture to our package. We will create a new directory
in our package called `inst/images` and put our cat picture in there,
calling it `hooper_kitten.png`.

```{r eval=FALSE}
dir.create('inst/images', recursive=TRUE)
```

And to add our cat picture:

```{r eval=FALSE}
download.file('https://github.com/seandavi/CuteCats/raw/main/inst/images/hooper_kitten.png', destfile='inst/images/hooper_kitten.png')
```

## Step 5: Document functions

Adding documentation to functions is really important for you, your future
self, and anyone you share your functions and package with. Doing so looks
intimidating at first, but once you do it a couple of times, it is 
pretty straightforward. The approach most people use these days is to document
functions using the [roxygen package](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html). 

Edit your `love_cats` function file to look like:

```{r eval=FALSE}
#' Do you love cats?
#'
#' This function allows you to express your love of cats.
#' 
#' @param love Do you love cats? Defaults to TRUE.
#' @keywords cats
#' @examples
#' love_cats()
#' @export
love_cats <- function(love=TRUE){
    if(love==TRUE){
        print("I love cats!")
        return(TRUE)
    }
    else {
        print("I am not a cool person.")
        return(FALSE)
    }
}

```

Save this file again. Now, to get the documentation from the comments in 
the code above to an actual R documentation file, run:

```{r eval=FALSE}
devtools::document()
```

Do the same for the `hooper_kitten` function:

```{r eval=FALSE}
#' Hooper Kitten picture
#' 
#' Displays the cutest kitten at Cold Spring Harbor
#' @examples
#' hooper_kitten()
#' @export
hooper_kitten <- function() {
    file_loc = system.file('images/hooper_kitten.png', package='CuteCats')
    pic = png::readPNG(file_loc)
    grid::grid.raster(img)
}
```

## Step 6: Try it out

The `devtools` package includes a nice function to quickly "load" the package
without having to install it, making it convenient and fast to make changes
to the package and test them out. 

```{r eval=FALSE}
devtools::load_all()
```

And try out the functions. 

## Step 7: Add a license

```{r eval=FALSE}
usethis::use_mit_license()
```

## Step 8: Build a pkgdown site

```{r eval=FALSE}
BiocManager::install('pkgdown')
```

Then, build the site:

```{r eval=FALSE}
pkgdown::build_site()
```

The site will end up in the `docs` directory. We will use that information
to tell GitHub where the website material is located.

## Step 9: Add everything to git

If you've cloned the package from git, no need to do this step. However, if 
you created the package from scratch, you'll want to do this to add 
version control to your package. 

On the terminal command line, switch into your package directory and then do:

```
git init
git add .
git commit -m 'Working package'
```

## Step 10-Infinity: Iterate and improve



